cubble An R Package for Structuring Spatio-temporal Data

### 2022-04-20 arxiv/cubble-arxiv.tex
work on section 3 the cubble package

Spatio-temporal data involves variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time intervals. This type of data can naturally be
considered as a data cube with three axes: spatial identifier, temporal
identifier, and variables. In this cubic formation, space only occupies
one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows three examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.001}

\caption{An illustration of different spatio-temporal data layouts: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.}\label{fig:illu-input}
\end{figure}

While incorporating both spatial and temporal information enriches the
data in the analysis, spatio-temporal data can sometimes be hard to work
with given its two observational units in the variables. The single
table format in Figure \ref{fig:illu-input} is convenient for
spatio-temporal operations that involve both types of variables,
however, it makes duplicates of spatial variables and is not ideal when
only looking at the spatial aspect of the data. The two tables format
allows separate operations on spatial and temporal variables while
analysts need to work with two data objects simultaneously and join them
together into the single table format for spatio-temporal operations.
Neither of the two formats is desirable at the same time for spatial,
temporal, and spatio-temporal operations. This issue implies analysts
need to constantly think about the data format and arrange variables
into the proper shape before the actual operations, which can be a
laborious step when switching between the two formats back and forth
several times during the analysis.

In the \proglang{R} community, many data objects have been proposed to
work with specific spatio-temporal tasks, while few can freely explore
the data. \pkg{spacetime} (\protect\hyperlink{ref-spacetime}{E. Pebesma
2012}) is an attempt for such a data object, but the spatial and
temporal classes it is built on have now been gradually replaced by more
recent implementations. The \pkg{stars}
(\protect\hyperlink{ref-stars}{E. Pebesma 2021}) package uses an array
to wrangle spatio-temporal data, but the dense array it uses can quickly
exhaust the memory for data in the case of irregular spatial grid. While
there has not yet been a data object that makes it easy to wrangle
spatio-temporal data, there is a strong need for such tools given the
ubiquitous spatial and temporal data collected in the society, i.e.,
climate observations from weather stations, air quality variables from
air monitoring stations. The demand for such tools can also come from
spatial data analysts wishing to incorporate temporal data in their
analysis or vice versa.

This paper presents a new \proglang{R} package, \pkg{cubble}, to
overcome the problem with organising variables in different levels when
working with spatio-temporal data. In the package a new data structure,
also called \code{cubble}, is proposed to organise spatial and temporal
variables as two forms of a single data object so that they can be
wrangled separately while synchronised as a whole piece. With this new
data object, analysts can spend less time writing codes to organise
spatial and temporal variables and focus more on the exploratory data
analysis itself. The software is available from the Comprehensive R
Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section 2 presents the main
design and functionality of \pkg{cubble}. Section 3 explains how cubble
deals with more advanced considerations including data with hierarchical
structure, data matching, how cubble fits with existing static and
interactive visualisation tools, and spatio-temporal data
transformation. Section 4 uses Australia weather station data and river
level data as examples to demonstrate the use of \pkg{cubble}. An
example on how \pkg{cubble} handles NetCDF data is also provided.
Section 5 concludes the paper.

\hypertarget{conceptual-framework-spatio-temporal-cube}{%
\section{Conceptual framework: spatio-temporal
cube}\label{conceptual-framework-spatio-temporal-cube}}

Multivariate spatio-temporal data can be conceptualised using a cubical
data model with reference to variables, time and space. This naturally
motivates the usage of array to represent and store spatio-temporal
data, as evident by satellite imageries, large climate models, among
others. The work by Lu, Appel, and Pebesma
(\protect\hyperlink{ref-lu_multidimensional_2018}{2018}) has discussed
the suitability of this representation through array algebra and
practical spatio-temporal tasks.

@@

\hypertarget{existing-work-and-new-challenges}{%
\section{Existing work and new
challenges}\label{existing-work-and-new-challenges}}

To represent spatio-temporal data, E. Pebesma
(\protect\hyperlink{ref-spacetime}{2012}) proposes four layouts: full
grid layout (STF), sparse grid layout (STS), irregular layout (STI), and
trajectory (STT). These layouts are implemented in the \pkg{spacetime}
package with underlying class \pkg{sp} (\protect\hyperlink{ref-sp}{E.
Pebesma and Bivand 2005}) and \pkg{xts}
(\protect\hyperlink{ref-xts}{Ryan and Ulrich 2020}) for handling spatial
and temporal variables. While these layouts and classes are still being
used and well-maintained today, they are not compatible with the
\pkg{tidyverse} ecosystem (\protect\hyperlink{ref-tidyverse}{Wickham et
al. 2019}), which has a huge influence on data analysis in \proglang{R},
including in the spatial and temporal communities. The package \pkg{sf}
(\protect\hyperlink{ref-sf}{E. J. Pebesma 2018}) uses data frame objects
to represent spatial data with an additional \code{geometry} column
dedicating to feature geometries
(\protect\hyperlink{ref-jr_herring_opengis_2011}{J.R. Herring 2011})
(e.g.~POINT, MULTIPOINT, POLYGON, MULTIPOLYGON etc). For temporal data,
\pkg{tsibble} (\protect\hyperlink{ref-tsibble}{Wang, Cook, and Hyndman
2020a}) builds on top of the \pkg{tibble}
(\protect\hyperlink{ref-tibble}{Müller and Wickham 2021}) object and
explicitly includes date and time in columns as variables, rather than
data attributes.

Underlying tidyverse is the concept of tidy data
(\protect\hyperlink{ref-tidydata}{Wickham 2014}), which considers how
variables should be structured for analysis. In the three formats
(time-wide, space-wide, and long) summarised in the E. Pebesma
(\protect\hyperlink{ref-spacetime}{2012}) paper, only the long form is
considered as tidy. The main advantages of it over the time-wide and
space-wide format is that the latter treat either time or location as
variable names (Section 3.1 in Wickham
(\protect\hyperlink{ref-tidydata}{2014})) rather than values of
variables, making them difficult to be wrangled. One problem with the
long format is that it can be inefficient to store feature geometries,
especially for large MULTIPOLYGON over long time period. This motivates
a new tidyverse compatible data object to work with spatio-temporal
data.

\textbf{two challenages: multiple data sources + large spatio-temporal
object} In additional to the shift into more tidyverse workflow for
working with data, spatio-temporal data have some unique challenges. The
first one being the split of variables in multiple tables. Often the
time, spatio-temporal data is not presented in a single data object, but
split into fragmented pieces from different sources. For example, map
data recorded the shape of a collection of area of interest is provided
by the government website; the geo-scientific data is recorded with
longitude and latitude coordinate of locations in the area; and temporal
variables indexed by both location and time need to be further queried
using location data. For spatio-temporal data analysis, merging all
these sources can be challenging and joining these tables can go wrong
because of the slightest unmatched of the linking variable from
different sources.

Another challenge when analysing spatio-temporal data is on the simple
feature representation of spatial data. While the simple feature is able
to handle large and complex spatial geometry and perform precise spatial
operations, data analysts usually do not have a large expectation on the
map accuracy and would prefer a simplified map with quick rendering for
visualisation. Also, the feature geometry column in \pkg{sf} wraps the
spatial coordinates in a list column, sometimes, analysts may like to
unpack them into longitude and latitude to work with. A smooth
transition with the coordinate columns and and simple feature geometry
column would satisfy analysts from both ends.

With some of the aforementioned issues recognised, the \pkg{stars}
package has been a new implementation to handle spatio-temporal data
conceptually as data cubes and implemented as arrays. The package
depends on the \pkg{sf} package and has allowed more access to tidyverse
verbs in the workflow. However, for vector spatio-temporal data, major
difficulties to work with \pkg{stars} is first to cast the variables
from multiple tables into a star object and then keep the operations in
the high-dimensional cube. Given that most table mentioned before
collect variables in a 2D table, analysts could benefit from a structure
to allow them work in 2D tables.

\hypertarget{the-cubble-package}{%
\section{The cubble package}\label{the-cubble-package}}

This section will present two formats that cubble uses to arrange
spatio-temporal data. The main cubble functions will then be introduced
to illustrate how to work with these two formats in short examples.
Lastly, a subsection will be dedicated to how existing packages, in
spatial and temporal analysis, fit in with cubble.

In cubble, a single data object can be arranged into two forms: nested
form and long form. Figure \ref{fig:illu-cubble} sketches the two forms
with the associated attributes. The decision on which form to use in
data wrangling is output-oriented, meaning analysts need to first think
about whether the output of an operation is identified only by the
spatial identifier, or a combination of spatial and temporal identifier.
The nested cubble is suitable for working with operations that are only
identified by the site and this type of operation can be a pure
manipulation of time invariant variables, or an operation that
summarises time varying variables into site. Underneath the nested form,
a cubble is built from a \code{rowwise_df} class where each site forms a
separate group. This structure simplifies the calculation that involves
temporal variables by avoiding the use of \code{purrr::map()} syntax
when working with list-column.

For those operations whose output involves both a spatial and temporal
identifier, long form should be used. The long form uses a
\code{grouped_df} class to forms all the time of a site as a group. Time
invariant variables are stored separately as an special attribute of the
long cubble. This design avoids repeating the spatial variables at each
time stamp while not dropping information from spatial variables.

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.002}

\caption{An illustration of the nested and long form in cubble. The nested form defines each site in a row and nests the time varying variables into a single column \code{ts}. The long form cubble uses \code{id} and \code{t} to identify each row and stores the time invariant variables as an attribute, \code{spatial}.}\label{fig:illu-cubble}
\end{figure}

\hypertarget{create-a-cubble-in-the-nested-form}{%
\subsection{Create a cubble in the nested
form}\label{create-a-cubble-in-the-nested-form}}

To use functions in the \texttt{cubble} package, an analyst will first
need to turn the data object into a \code{cubble} class.
\code{as_cubble()} does this through supplying the three key components:
\code{key} as the spatial identifier; \code{index} as the temporal
identifier; and a vector of \code{coords} in the order of longitude and
latitude. The use of \code{key} and \code{index} follows the design in
the \pkg{tsibble} package and the cubble created by default is in the
nested form.

Before the formal examples in Section \ref{examples}, each function
introduced in this section will be accompanied by a small example. The
data used for the small examples is a subset of a larger climate weather
station data used in the formal examples and has been simplified to
include only five weather stations. It contains spatial information of
each station: station id, latitude, longitude, elevation, station name
and World Meteorology Organisation ID and also daily temporal
information of date, maximum and minimum temperature and precipitation
for 2020. \code{climate_flat} stores this data in format 1) in Figure
\ref{fig:illu-input} and the code below creates a cubble out of
\code{climate_flat} with \code{id} as the key, \code{date} as the index,
and \code{c(long, lat)} as the coordinates:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_nested }\OtherTok{\textless{}{-}}\NormalTok{ cubble}\SpecialCharTok{::}\NormalTok{climate\_flat }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date, }\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}
\NormalTok{cubble\_nested}
\end{Highlighting}
\end{Shaded}

@@

There are a few information in the cubble header: the name of the
\code{key} variable: \code{id} and its unique number: 5, the bounding
box: \code{[115.97, -32.94, 133.55, -12.42]} , and also the name of
variable nested in the \code{ts} column with its type:
\code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

\hypertarget{stretch-a-nested-cubble-into-the-long-form}{%
\subsection{Stretch a nested cubble into the long
form}\label{stretch-a-nested-cubble-into-the-long-form}}

The nested format is convenient for those operations whose output does
not contain a time dimension. For those outputs that are
cross-identified by the spatial and temporal identifier, a long cubble
needs to be used. The function \code{stretch()} is designed to switch
the cubble from the nested form to the long form. This code shows how to
switch the nested cubble just created into its long form:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_long }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_nested }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_temporal}\NormalTok{()}
\NormalTok{cubble\_long}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{tamp-a-long-cubble-back-to-the-nested-form}{%
\subsection{Tamp a long cubble back to the nested
form}\label{tamp-a-long-cubble-back-to-the-nested-form}}

Manipulation on the spatial and temporal dimension can be an iterative
process and analysts may need to go back and forth between the nested
and long cubble. \code{tamp()}, an inverse of \code{stretch()}, switches
a long cubble to a nested cubble:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_back }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_long }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_spatial}\NormalTok{()}
\NormalTok{cubble\_back}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{migrate-spatial-variables-to-a-long-cubble}{%
\subsection{Migrate spatial variables to a long
cubble}\label{migrate-spatial-variables-to-a-long-cubble}}

Sometimes, analysts may need to apply some variable transformation that
involves both the spatial and temporal variable. An example of this is
the transformation of temporal variables into the spatial dimension in
glyph maps (which will be elaborated in section
\ref{st_transformation}). Cubble allows this operation through
\code{unfold()}, which moves the specified spatial variables into the
long cubble:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_mig }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_long }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{unfold}\NormalTok{(long, lat)}
\NormalTok{cubble\_mig}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_mig }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_spatial}\NormalTok{() }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_temporal}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{compatibility-with-existing-packages}{%
\subsection{Compatibility with existing
packages}\label{compatibility-with-existing-packages}}

@@

\code{tsibble} is a temporal data structure that uses \code{index} and
\code{key} to identify the time and different series. \code{cubble} can
be seen as a natural extension of \code{tsibble} for spatio-temporal
data with an additional coordinates component and using two forms to
arrange variables. This makes it easy to cast a \code{tsibble} into a
\code{cubble} as only the \code{coords} argument needs to be supplied:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# example with a tsibble created from climate\_flat}
\NormalTok{raw }\OtherTok{\textless{}{-}}\NormalTok{ climate\_flat }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
\NormalTok{  tsibble}\SpecialCharTok{::}\FunctionTok{as\_tsibble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date)}

\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{  raw }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
\NormalTok{  cubble}\SpecialCharTok{::}\FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}
\NormalTok{dt}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# add station{-}based features in the nested form.}
\NormalTok{dt }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{mutate}\NormalTok{(fabletools}\SpecialCharTok{::}\FunctionTok{features}\NormalTok{(ts, tmax, }\FunctionTok{list}\NormalTok{(}\AttributeTok{tmax\_mean =}\NormalTok{ mean, }\AttributeTok{tmax\_var =}\NormalTok{ var)))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(sf)}
\CommentTok{\# create a cubble}
\NormalTok{cb }\OtherTok{\textless{}{-}}\NormalTok{ climate\_flat }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
\NormalTok{  cubble}\SpecialCharTok{::}\FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date, }\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}

@@

\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{ cb }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{mutate}\NormalTok{(}
    \CommentTok{\# create \textasciigrave{}sfc\textasciigrave{} column based on long and lat}
    \AttributeTok{ll =} \FunctionTok{st\_sfc}\NormalTok{(}
\NormalTok{      purrr}\SpecialCharTok{::}\FunctionTok{map2}\NormalTok{(long, lat, }\SpecialCharTok{\textasciitilde{}}\FunctionTok{st\_point}\NormalTok{(}\FunctionTok{c}\NormalTok{(.x, .y))),}
      \AttributeTok{crs =} \FunctionTok{st\_crs}\NormalTok{(aus)),}

@@

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and may have multiple variables measured at each spatial and temporal value. The analysis of spatio-temporal data can fix the time to explore the spatial distribution, fix the spatial location to explore temporal trends, or dynamically explore the space and time simultaneously. For these tasks to be carried out easily, there is a need for a spatio-temporal data representation that underpins various operations.

The \emph{Representing data} section of the SpatioTemproal CRAN task view (\protect\hyperlink{ref-ctvspatiotemporal}{Edzer Pebesma 2022}) lists existing spatio-temporal data representation. Among them, E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}) proposes four layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} (\protect\hyperlink{ref-spacetime}{E. Pebesma 2012}) package with the underlying spatial and temporal class: \pkg{sp} (\protect\hyperlink{ref-sp}{E. Pebesma and Bivand 2005}) and \pkg{xts} (\protect\hyperlink{ref-xts}{Ryan and Ulrich 2020}). The \pkg{stars} (\protect\hyperlink{ref-stars}{E. Pebesma 2021}) package has been a recent implementation to handle both raster and vector spatio-temporal data using array. It depends on the newer spatial package \pkg{sf} (\protect\hyperlink{ref-sf}{E. J. Pebesma 2018}) and interfaces with the widely-used tidyverse verbs.

{[}till here{]} One issue with wrangling spatio-temporal data is that the raw data does not naturally structured in any one of the representation above. Also, in data collection, spatio-temporal data is less often collected as a whole piece but fragmented spatial and temporal data for analysts to ensemble. For example, map data records the shape of a collection of area of interest is provided by official government website; geo-scientific data records the longitude and latitude coordinate of locations in the area along with other meta data relates to the location; and temporal variables indexed by both location and time. For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.

Another issue is how the concept of tidy data (\protect\hyperlink{ref-tidydata}{Wickham 2014}) should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e.~\pkg{dplyr} and \pkg{ggplot2} for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period. This poses the question on how to arrange spatial and temporal variables that makes analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble}, to {[}overcome the problem{]} with organising variables in different levels when working with spatio-temporal data. The data structure focuses on connecting to the upstream data collection and is suitable for temporal data collected at the fixed location over time. Among the four spacetime layouts in E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}), it can be full grid layout, sparse grid layout, or irregular layout, but trajectory is not within the scope of this work. In the package a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. With this new data object, analysts can spend less time writing codes to organise spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section \ref{cube} discusses the spatio-temporal cube to conceptualise spatio-temporal data. Section \ref{cubble} presents the main design and functionality of \pkg{cubble}. Section \ref{others} explains how cubble deals with more advanced considerations including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \ref{examples} uses Australia weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example on how \pkg{cubble} handles NetCDF data is also provided. Section \ref{conclude} concludes the paper.

\hypertarget{cube}{%
\section{Conceptual framework: spatio-temporal cube}\label{cube}}

Spatio-temporal data can be conceptualised using a cubical data model with reference to variables, time and space. This naturally motivates the usage of array to represent and store spatio-temporal data, as evident by satellite imageries, large climate models, among others. The work by Lu, Appel, and Pebesma (\protect\hyperlink{ref-lu_multidimensional_2018}{2018}) has discussed the suitability of this representation through array algebra and practical spatio-temporal tasks. \textcolor{magenta}{SZ: is this too direct?}

@@

\hypertarget{cubble}{%
\section{The cubble package}\label{cubble}}

This section will first introduce the core functions in the cubble class: \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()} from subsection \ref{create} to \ref{unfold}. The next subsection addresses why bother a new class before subsection \ref{compact} finishing on the compatibility of cubble with existing packages in spatial and temporal analysis.

Each core function is introduced with a short example accompanied and the data \texttt{climate\_flat} is used throughout this section. \texttt{climate\_flat} contains five weather stations in Australia with spatial information of each station: station id, latitude, longitude, elevation, station name and World Meteorology Organisation ID and also daily temporal information of date, maximum and minimum temperature and precipitation for 2020. Below prints out the first five rows of \texttt{climate\_flat}:

\begin{verbatim}
## # A tibble: 1,830 x 10
##   id            lat  long  elev name         wmo_id date        prcp  tmax  tmin
##   <chr>       <dbl> <dbl> <dbl> <chr>         <dbl> <date>     <dbl> <dbl> <dbl>
## 1 ASN00009021 -31.9  116.  15.4 perth airpo~  94610 2020-01-01     0  31.9  15.3
## 2 ASN00009021 -31.9  116.  15.4 perth airpo~  94610 2020-01-02     0  24.9  16.4
## 3 ASN00009021 -31.9  116.  15.4 perth airpo~  94610 2020-01-03     6  23.2  13  
## 4 ASN00009021 -31.9  116.  15.4 perth airpo~  94610 2020-01-04     0  28.4  12.4
## 5 ASN00009021 -31.9  116.  15.4 perth airpo~  94610 2020-01-05     0  35.3  11.6
## # ... with 1,825 more rows
\end{verbatim}

\hypertarget{create}{%
\subsection{Create a cubble in the nested form}\label{create}}

As described in the introduction, spatio-temporal data can come in various formats and shapes. A cubble can be created from various raw data, including tibble and its variates multiple tables (an example provided in section \ref{covid}), and netcdf (detailed in section \ref{netcdf}). The function \code{as_cubble()} is used to create a cubble with three additional components: \code{key} as the spatial identifier; \code{index} as the temporal identifier; and a vector of \code{coords} in the order of longitude and latitude. Argument \code{key} and \code{index} follow the wording in \pkg{tsibble} to define temporal order and multiple series while \code{coords} is a space specific argument to locate each site in two columns. The code below creates a cubble out of \code{climate_flat} with \code{id} as the key, \code{date} as the index, and \code{c(long, lat)} as the coordinates:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_nested }\OtherTok{\textless{}{-}}\NormalTok{ climate\_flat }\SpecialCharTok{|\textgreater{}}
  \FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date, }\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}
\NormalTok{cubble\_nested}
\end{Highlighting}
\end{Shaded}

@@

The cubble header provides some information about this data. \code{id} is the variable name to identify each location and there are five unique locations. The bounding box is \code{[115.97, -32.94, 133.55, -12.42]} and the names of variable nested in the \code{ts} column, along with its type, are \code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

The created cubble is built from a \code{rowwise_df} class where each row forms a group and all the temporal variables are nested in a list column, hence it is also called the nested cubble. The rowwise structure makes it simpler to calculate on the list using \code{mutate()} and avoid the \code{map()} when working with list column. For example calculating the number of raining day can be done by:

\begin{verbatim}
## # cubble:   id [5]: nested form
## # bbox:     [115.97, -32.94, 133.55, -12.42]
## # temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
##   id            lat  long  elev name           wmo_id ts       rain_day
##   <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>      <int>
## 1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble>      104
## 2 ASN00010311 -31.9  117. 179   york            94623 <tibble>       89
## 3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble>       90
## 4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble>      106
## 5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble>       63
\end{verbatim}

\hypertarget{switch-from-the-nested-cubble-into-the-long-form}{%
\subsection{Switch from the nested cubble into the long form}\label{switch-from-the-nested-cubble-into-the-long-form}}

The nested format is convenient for those operations whose output does not contain a time dimension. For those outputs that are cross-identified by the spatial and temporal identifier, a long cubble needs to be used. The function \code{face_temporal()} switches the cubble from the nested form to the long form and the first row in Figure \ref{fig:face} sketches the function in a diagram. This code shows how to switch the cubble just created into its long form:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_long }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_nested }\SpecialCharTok{|\textgreater{}} \FunctionTok{face\_temporal}\NormalTok{()}
\NormalTok{cubble\_long}
\end{Highlighting}
\end{Shaded}

@@

The long form is built form class \code{groupped_df} where all the observations from the same sites forms a group.

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.008}

\caption{An illustration of function \code{face\_temporal} and \code{face\_spatial} in cubble. In the first row, \code{face\_temporal} switches a cubble from the nested form into the long form and the focus has switched from the spatial aspect (the side face) to the temporal aspect (the front face). In the second row, \code{face\_spatial} switches a cubble back to the nested form from te long form and shift focus back to the spatial aspect.}\label{fig:face}
\end{figure}

\hypertarget{switch-a-long-cubble-back-to-the-nested-form}{%
\subsection{Switch a long cubble back to the nested form}\label{switch-a-long-cubble-back-to-the-nested-form}}

Manipulation on the spatial and temporal dimension can be an iterative process and analysts may need to go back and forth between the nested and long cubble. \code{face_spatial()}, the inverse of \code{face_temporal()}, switches the long cubble back to the nested form, as sketched in the second row in Figure \ref{fig:face}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_back }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_long }\SpecialCharTok{|\textgreater{}} \FunctionTok{face\_spatial}\NormalTok{()}
\NormalTok{cubble\_back}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{unfold}{%
\subsection{Flatten spatial variables into the long cubble}\label{unfold}}

Sometimes, analysts may need to apply some variable transformation that involves both the spatial and temporal variable. An example of this is the transformation of temporal variables into the spatial dimension in glyph maps, which will be elaborated in section \ref{st_transformation}. This type of operations can be seen as flattening, or \emph{unfolding}, the cube into a 2D data frame. Here the function \code{unfold()} moves the spatial variables \code{long} and \code{lat} into the long cubble:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_unfold }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_long }\SpecialCharTok{|\textgreater{}} \FunctionTok{unfold}\NormalTok{(long, lat)}
\NormalTok{cubble\_unfold}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_unfold }\SpecialCharTok{|\textgreater{}} \FunctionTok{face\_spatial}\NormalTok{() }\SpecialCharTok{|\textgreater{}} \FunctionTok{face\_temporal}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

@@

\subsection[Why not just use]{Why not just use \code{dplyr::nest\_by()}}

It may naturally raise the question that why bother a new data structure since the nested form can already be created by the \pkg{dplyr} function \code{nest_by()}? While \code{nest_by()} and \code{unnest()} can mimic the nested form in cubble, cubble can be seen as an attempt to arrange variables from different observational units into a single object, specifically for spatio-temporal data. While \texttt{nest\_by()} solves this problem on the spatial side, it doesn't solve the temporal side. For temporal data, a combined representation with spatial variables is useful for spatio-temporal operations but its duplication can be a burden to operate on. Also, spatial variables are dropped when summarised and you will need to either specified all of them in the \code{group_by()} before \code{summarise}, or add a joining after \code{summarise()}.

\hypertarget{compact}{%
\subsection{Compatibility with existing packages}\label{compact}}

@@

\code{tsibble} is a temporal data structure that uses \code{index} and \code{key} to identify the time and different series. \code{cubble} can be seen as following the same vein as \code{tsibble} for spatio-temporal data. This makes it easy to cast a \code{tsibble} into a \code{cubble} as only the \code{coords} argument needs to be supplied:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# example with a tsibble created from climate\_flat}
\NormalTok{raw }\OtherTok{\textless{}{-}}\NormalTok{ climate\_flat }\SpecialCharTok{|\textgreater{}}\NormalTok{ tsibble}\SpecialCharTok{::}\FunctionTok{as\_tsibble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date)}
\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{  raw }\SpecialCharTok{|\textgreater{}}\NormalTok{ cubble}\SpecialCharTok{::}\FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}
\NormalTok{dt}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# add station{-}based features in the nested form.}
\NormalTok{dt }\SpecialCharTok{|\textgreater{}} \FunctionTok{mutate}\NormalTok{(fabletools}\SpecialCharTok{::}\FunctionTok{features}\NormalTok{(ts, tmax, }\FunctionTok{list}\NormalTok{(}\AttributeTok{tmax\_mean =}\NormalTok{ mean, }\AttributeTok{tmax\_var =}\NormalTok{ var)))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(sf)}
\CommentTok{\# create a cubble}
\NormalTok{cb }\OtherTok{\textless{}{-}}\NormalTok{ climate\_flat }\SpecialCharTok{|\textgreater{}}
\NormalTok{  cubble}\SpecialCharTok{::}\FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date, }\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}

@@

\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{ cb }\SpecialCharTok{|\textgreater{}}
  \FunctionTok{mutate}\NormalTok{(}
    \CommentTok{\# create \textasciigrave{}sfc\textasciigrave{} column based on long and lat}
    \AttributeTok{ll =} \FunctionTok{st\_sfc}\NormalTok{(}
\NormalTok{      purrr}\SpecialCharTok{::}\FunctionTok{map2}\NormalTok{(long, lat, }\SpecialCharTok{\textasciitilde{}}\FunctionTok{st\_point}\NormalTok{(}\FunctionTok{c}\NormalTok{(.x, .y))),}
      \AttributeTok{crs =} \FunctionTok{st\_crs}\NormalTok{(aus)),}

@@

\hypertarget{covid}{%
\subsection{Victoria covid data}\label{covid}}

This example illustrate casting different sources into a cubble object. Victoria government of Australia provides daily COVID case number by date, source, and local government area (LGA). This data can be useful for visualising COVID spread when combined with LGA map information, which is available from Australian Bureau of Statistics. Variable \code{covid} and \code{lga} show the first 5 rows of these two data:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{covid }\SpecialCharTok{|\textgreater{}} \FunctionTok{head}\NormalTok{(}\DecValTok{5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}
